```{r ttest_example}
# Create example data for comparing two groups
set.seed(123)                                # Ensure reproducibility
control <- rnorm(30,                         # Generate control group data
                mean = 100,                  # Set control group mean
                sd = 15)                     # Set control group variation
treatment <- rnorm(30,                       # Generate treatment group data
                  mean = 115,                # Set treatment group mean
                  sd = 15)                   # Match control group variation

# Perform statistical analysis
t_results <- t.test(control, treatment)      # Compare groups using t-test

# Create organized results table
tibble(
  Measure = c("Mean Difference",             # Key statistical measures
              "t-statistic", 
              "p-value", 
              "95% CI"),
  Value = c(
    round(diff(c(mean(control),             # Calculate mean difference
                 mean(treatment))), 2),
    round(t_results$statistic, 2),          # Extract t-statistic
    format.pval(t_results$p.value, 3),      # Format p-value nicely
    paste(round(t_results$conf.int, 2),     # Format confidence interval
          collapse = " to ")
  )
) %>%
  knitr::kable(caption = "Statistical Test Results")

# Prepare data for visualization
test_data <- tibble(
  value = c(control, treatment),            # Combine all measurements
  group = rep(c("Control", "Treatment"),    # Label each group
             each = 30)
)

# Create informative visualizations
# 1. Density plot to show distribution shapes
density_plot <- ggplot(test_data, 
                      aes(x = value, 
                          fill = group)) +
  geom_density(alpha = 0.5) +               # Semi-transparent density curves
  labs(
    title = "Group Comparison: Density Plot",
    subtitle = paste("p-value =",           # Show statistical significance
                    format.pval(t_results$p.value, 3)),
    x = "Measurement Value",
    y = "Density"
  ) +
  theme_minimal()

# 2. Box plot with data points for detailed view
box_plot <- ggplot(test_data, 
                  aes(x = group, 
                      y = value, 
                      fill = group)) +
  geom_boxplot(alpha = 0.5) +              # Show distribution summary
  geom_jitter(width = 0.2,                 # Add individual points
              alpha = 0.5) +               # Semi-transparent points
  labs(
    title = "Group Comparison: Box Plot",
    subtitle = paste("Mean Difference =",   # Show effect size
                    round(mean(treatment) - mean(control), 2)),
    x = "Group",
    y = "Measurement Value"
  ) +
  theme_minimal()

# Display both plots side by side
grid.arrange(density_plot, box_plot, ncol = 2)

# Check statistical assumptions
# 1. Normality check using Q-Q plots
qq_plots <- map(list(Control = control,     # Create Q-Q plot for each group
                    Treatment = treatment),
                ~{
                  ggplot(data.frame(x = .x), 
                         aes(sample = x)) +
                    stat_qq() +             # Add Q-Q points
                    stat_qq_line() +        # Add reference line
                    labs(title = paste("Q-Q Plot:", 
                                     deparse(substitute(.x)))) +
                    theme_minimal()
                })

# Display Q-Q plots side by side
grid.arrange(grobs = qq_plots, ncol = 2)

# 2. Variance equality check
var_test <- var.test(control, treatment)    # Test for equal variances

# Create variance test results table
tibble(
  Measure = c("F statistic",               # Key measures for variance test
              "p-value",
              "Ratio of variances"),
  Value = c(
    round(var_test$statistic, 3),          # F-statistic
    format.pval(var_test$p.value, 3),      # p-value
    round(var(treatment)/var(control), 3)   # Actual ratio of variances
  )
) %>%
  knitr::kable(caption = "Variance Test Results")

# 3. Independence check
independence_plot <- tibble(
  Order = 1:30,                            # Observation order
  Control = control,                       # Control group values
  Treatment = treatment                    # Treatment group values
) %>%
  pivot_longer(cols = c(Control, Treatment),
              names_to = "Group",
              values_to = "Value") %>%
  ggplot(aes(x = Order, 
             y = Value, 
             color = Group)) +
  geom_point() +                          # Show individual measurements
  geom_line(alpha = 0.5) +               # Connect points to show patterns
  facet_wrap(~Group) +                   # Separate plots by group
  labs(
    title = "Independence Check",
    subtitle = "Looking for patterns in measurement order",
    x = "Observation Order",
    y = "Measurement Value"
  ) +
  theme_minimal()

# Display independence check plot
independence_plot

```{r scatter_plot}
# Analyze relationships in iris dataset
iris_analysis <- iris %>%
  # Calculate correlation between sepal and petal length
  summarise(
    correlation = cor(Sepal.Length, Petal.Length),  # Pearson correlation
    p_value = cor.test(Sepal.Length,               # Test significance
                      Petal.Length)$p.value
  )

# Create comprehensive scatter plot
ggplot(iris, 
       aes(x = Sepal.Length,                       # X-axis variable
           y = Petal.Length,                       # Y-axis variable
           color = Species)) +                     # Color by species
  # Add individual data points
  geom_point(alpha = 0.7) +                       # Semi-transparent points
  # Add regression lines for each species
  geom_smooth(method = "lm",                      # Linear regression
             se = TRUE,                           # Show confidence interval
             alpha = 0.2) +                       # Transparent confidence bands
  # Add informative labels
  labs(
    title = "Relationship between Sepal and Petal Length",
    subtitle = paste("Overall correlation =",      # Show correlation
                    round(iris_analysis$correlation, 3)),
    x = "Sepal Length (cm)",
    y = "Petal Length (cm)"
  ) +
  # Add regression equations
  stat_smooth(method = "lm",                      # Add equation for each species
             geom = "text",
             aes(label = paste("R² =", 
                             round(stat(r.squared), 3))),
             hjust = 0,
             vjust = -1) +
  theme_minimal() +
  theme(legend.position = "bottom")               # Move legend to bottom

## Practice Exercise Solutions

### Solution 1: Basic Data Exploration
```{r solution1}
# Comprehensive summary of mtcars dataset
car_summary <- mtcars %>%
  summarise(across(
    everything(),                                 # Apply to all variables
    list(
      mean = ~mean(.x),                          # Calculate mean
      sd = ~sd(.x),                              # Calculate std deviation
      min = ~min(.x),                            # Find minimum
      max = ~max(.x)                             # Find maximum
    ),
    .names = "{.col}_{.fn}"                      # Name format: var_stat
  )) %>%
  pivot_longer(everything(),                      # Convert to long format
               names_to = c("variable", "stat"),  # Split names
               names_sep = "_") %>%
  pivot_wider(names_from = stat,                 # Reshape for presentation
              values_from = value)

# Display formatted summary table
knitr::kable(car_summary,
             caption = "Summary Statistics for Car Features",
             digits = 2)                          # Round to 2 decimals

# Create distribution visualization
ggplot(mtcars, aes(x = wt)) +
  # Add histogram with density scaling
  geom_histogram(aes(y = after_stat(density)),    # Modern syntax for density
                fill = "skyblue",
                color = "white",
                bins = 30) +
  # Add density curve
  geom_density(color = "red",                     # Add smooth curve
               linewidth = 1) +
  # Add mean line
  geom_vline(aes(xintercept = mean(wt)),         # Show average
             color = "darkred",
             linetype = "dashed") +
  # Add informative labels
  labs(
    title = "Distribution of Car Weights",
    subtitle = paste("Mean =", round(mean(mtcars$wt), 2),
                    "±", round(sd(mtcars$wt), 2), "thousand lbs"),
    x = "Weight (thousand lbs)",
    y = "Density"
  ) +
  theme_minimal()
```

### Solution 2: Group Comparisons
```{r solution2}
# Prepare data for analysis
cylinder_stats <- mtcars %>%
  group_by(cyl) %>%                              # Group by cylinder count
  summarise(
    n = n(),                                     # Count cars per group
    mean_mpg = mean(mpg),                        # Calculate mean MPG
    sd_mpg = sd(mpg),                           # Calculate std deviation
    se_mpg = sd_mpg/sqrt(n)                     # Calculate std error
  )

# Create comprehensive visualization
ggplot(mtcars, aes(x = factor(cyl),             # Convert cylinders to factor
                   y = mpg,
                   fill = factor(cyl))) +
  # Add violin plot for distribution shape
  geom_violin(alpha = 0.3) +                     # Show distribution shape
  # Add box plot for summary statistics
  geom_boxplot(width = 0.2,                      # Narrow box width
               alpha = 0.7) +                     # Semi-transparent
  # Add individual points
  geom_jitter(width = 0.1,                       # Minimal horizontal jitter
              alpha = 0.5) +                     # Semi-transparent points
  # Add mean ± SE
  geom_pointrange(data = cylinder_stats,         # Add error bars
                 aes(y = mean_mpg,
                     ymin = mean_mpg - se_mpg,
                     ymax = mean_mpg + se_mpg),
                 color = "darkred") +
  # Add informative labels
  labs(
    title = "Car Mileage by Number of Cylinders",
    subtitle = "Showing distribution, box plot, and mean ± SE",
    x = "Number of Cylinders",
    y = "Miles Per Gallon",
    fill = "Cylinders"
  ) +
  theme_minimal()

# Perform statistical analysis
# 1. ANOVA test
anova_result <- aov(mpg ~ factor(cyl), data = mtcars)
# 2. Tukey's HSD for pairwise comparisons
tukey_result <- TukeyHSD(anova_result)

# Display ANOVA results
knitr::kable(broom::tidy(anova_result),
             caption = "ANOVA Results",
             digits = 3)

# Display Tukey results
knitr::kable(broom::tidy(tukey_result),
             caption = "Tukey's HSD Pairwise Comparisons",
             digits = 3)
```

### Solution 3: Relationship Analysis
```{r solution3}
# Perform correlation analysis
cor_result <- cor.test(mtcars$wt, mtcars$mpg)    # Test correlation

# Create comprehensive scatter plot
ggplot(mtcars, aes(x = wt, y = mpg)) +
  # Add individual points
  geom_point(aes(size = disp),                    # Size points by displacement
             alpha = 0.6) +                       # Semi-transparent
  # Add regression line with confidence interval
  geom_smooth(method = "lm",                      # Linear regression
             color = "red",
             alpha = 0.2) +                       # Transparent confidence band
  # Add correlation information
  annotate("text",
           x = min(mtcars$wt),                    # Position at left
           y = max(mtcars$mpg),                   # Position at top
           label = paste("r =",                    # Show correlation
                        round(cor_result$estimate, 3),
                        "\np <", format.pval(cor_result$p.value, 3)),
           hjust = 0,                             # Left-align text
           vjust = 1) +                           # Top-align text
  # Add informative labels
  labs(
    title = "Relationship between Weight and Mileage",
    subtitle = "Point size indicates engine displacement",
    x = "Weight (thousand lbs)",
    y = "Miles Per Gallon",
    size = "Displacement"
  ) +
  theme_minimal()

# Create summary statistics table
model <- lm(mpg ~ wt, data = mtcars)             # Fit linear model
model_summary <- broom::tidy(model)              # Extract model statistics

# Display regression results
knitr::kable(model_summary,
             caption = "Linear Regression Results",
             digits = 3)
``` 