---
title: "Introduction to Loops and Apply Functions in R"
author: "Eren Ada"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: united
---

```{r setup, include=FALSE}
# Set global options for all code chunks in this R Markdown document
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction to Loops and Apply Functions

In R, we often need to perform the same operation multiple times. We can do this using:
1. Loops (for, while)
2. Apply family functions (apply, lapply, sapply, mapply)

Let's learn both approaches with simple examples!

## Part 1: Loops in R

### For Loops

```{r}
# Example 1: Basic for loop that iterates from 1 to 5
for (i in 1:5) {
  # paste() combines strings, converting numbers to text if needed
  # This will print "This is iteration number X" for each value of i
  print(paste("This is iteration number", i))
}

# Example 2: Loop through elements in a vector
# Create a vector of fruit names
fruits <- c("apple", "banana", "orange", "grape")
# Loop through each fruit in the vector
for (fruit in fruits) {
  # Print a message for each fruit
  print(paste("I like", fruit))
}
```

### Practical Examples with For Loops

```{r}
# Example 1: Calculate squares of numbers
# Create a sequence of numbers from 1 to 5
numbers <- 1:5
# Create an empty numeric vector to store results (pre-allocation for efficiency)
squares <- numeric(length(numbers))  

# Loop through each number and calculate its square
for (i in 1:length(numbers)) {
  # Store the square of each number in the pre-allocated vector
  squares[i] <- numbers[i]^2
}
# Display the results
print(squares)

# Example 2: Working with a data frame
# Create a sample data frame with student information
students <- data.frame(
  name = c("Alice", "Bob", "Charlie"),    # Student names
  score1 = c(85, 92, 78),                 # First test scores
  score2 = c(88, 85, 90)                  # Second test scores
)

# Calculate and print average score for each student
for (i in 1:nrow(students)) {
  # Calculate mean of both scores for current student
  avg <- mean(c(students$score1[i], students$score2[i]))
  # Print formatted message with student name and average
  print(paste(students$name[i], "has an average score of", avg))
}
```

### While Loops

```{r}
# Example 1: Basic while loop
# Initialize counter
counter <- 1
# Continue loop as long as counter is less than or equal to 5
while (counter <= 5) {
  # Print current counter value
  print(paste("Counter is", counter))
  # Increment counter
  counter <- counter + 1
}

# Example 2: Sum numbers until exceeding 100
# Initialize variables
sum <- 0    # Running sum
count <- 0  # Number counter
# Continue adding numbers until sum reaches or exceeds 100
while (sum < 100) {
  count <- count + 1              # Increment counter
  sum <- sum + count             # Add current number to sum
  # Print progress message
  print(paste("Added", count, "Sum is now", sum))
}
```

### Nested Loops

```{r}
# Create a multiplication table using nested loops
for (i in 1:3) {                     # Outer loop: first number
  for (j in 1:3) {                   # Inner loop: second number
    # Calculate and print multiplication result
    print(paste(i, "x", j, "=", i*j))
  }
  print("---")  # Print separator between numbers
}
```

## Part 2: Apply Functions

### lapply(): List Apply

```{r}
# Create a list of numeric vectors
numbers <- list(a = 1:3, b = 4:6, c = 7:9)
# Apply sum function to each element of the list
lapply(numbers, sum)

# Create a custom function to double values
double_it <- function(x) {
  return(x * 2)
}
# Apply custom function to each element
lapply(numbers, double_it)

# Use anonymous function to triple values
lapply(numbers, function(x) x * 3)
```

### sapply(): Simplified Apply

```{r}
# Create a vector of numbers
numbers <- 1:5

# Compare lapply and sapply outputs
# lapply returns a list
lapply(numbers, sqrt)
# sapply returns a simplified structure (vector in this case)
sapply(numbers, sqrt)

# Example with character string lengths
words <- c("cat", "dog", "elephant", "rhinoceros")
# Calculate length of each word
sapply(words, nchar)
```

### apply(): Matrix or Array Apply

```{r}
# Create a 3x3 matrix with numbers 1-9
mat <- matrix(1:9, nrow = 3)
print(mat)

# Calculate mean of each row (MARGIN = 1)
apply(mat, 1, mean)

# Calculate mean of each column (MARGIN = 2)
apply(mat, 2, mean)

# Create a grade matrix for multiple students and subjects
grades <- matrix(c(
  85, 90, 88,  # Student 1's grades
  92, 85, 87,  # Student 2's grades
  78, 83, 85   # Student 3's grades
), nrow = 3, byrow = TRUE)

# Add column and row names for clarity
colnames(grades) <- c("Math", "Science", "History")
rownames(grades) <- c("Alice", "Bob", "Charlie")

print(grades)

# Calculate average grade for each student (row means)
student_averages <- apply(grades, 1, mean)
print(student_averages)

# Calculate average grade for each subject (column means)
subject_averages <- apply(grades, 2, mean)
print(subject_averages)
```

### mapply(): Multivariate Apply

```{r}
# Repeat letters different numbers of times
mapply(rep, times = 1:4, x = letters[1:4])

# Create a custom function for generating sequences
create_sequence <- function(start, end) {
  return(start:end)
}

# Generate multiple sequences with different start/end points
mapply(create_sequence, 
       start = c(1, 2, 3), 
       end = c(3, 4, 5))
```

## Practical Examples: Loops vs Apply Functions

Let's compare loops and apply functions for the same tasks:

```{r}
# Task: Calculate the square of each number
numbers <- 1:5

# Using a for loop
squares_loop <- numeric(length(numbers))
for (i in 1:length(numbers)) {
  squares_loop[i] <- numbers[i]^2
}

# Using sapply
squares_sapply <- sapply(numbers, function(x) x^2)

# Compare results
print(squares_loop)
print(squares_sapply)

# Task: Process a data frame
df <- data.frame(
  name = c("Alice", "Bob", "Charlie"),
  age = c(25, 30, 35),
  score = c(88, 92, 85)
)

# Using a for loop
for (i in 1:nrow(df)) {
  if (df$score[i] >= 90) {
    print(paste(df$name[i], "got an A!"))
  }
}

# Using sapply with an anonymous function
high_scorers <- sapply(1:nrow(df), function(i) {
  if (df$score[i] >= 90) return(paste(df$name[i], "got an A!"))
})
# Remove NULL results
high_scorers[!sapply(high_scorers, is.null)]
```

## When to Use What?

1. **Use For Loops When**:
   - The operation is complex
   - You need to modify the original data
   - You need more control over the iteration process
   - The code needs to be very clear for beginners

2. **Use Apply Functions When**:
   - The operation is simple
   - You're working with lists, vectors, or matrices
   - You want more concise code
   - Performance is important

## Practice Exercises

Try these exercises to practice:

```{r}
# Exercise 1: Generate Fibonacci sequence
# Initialize a vector to store the first 10 Fibonacci numbers
fibonacci <- numeric(10)
# Set the first two numbers of the sequence
fibonacci[1] <- 1    # First Fibonacci number
fibonacci[2] <- 1    # Second Fibonacci number

# Calculate the rest of the sequence using a for loop
for (i in 3:10) {
  # Each number is the sum of the previous two numbers
  fibonacci[i] <- fibonacci[i-1] + fibonacci[i-2]
}
# Display the complete sequence
print(fibonacci)

# Exercise 2: Calculate word lengths using sapply
# Create a vector of words to analyze
sentence <- c("The", "quick", "brown", "fox", "jumps")
# Use sapply to count characters in each word
word_lengths <- sapply(sentence, nchar)
# Display the length of each word
print(word_lengths)

# Exercise 3: Find maximum values in matrix rows
# Create a matrix with 4 rows and 5 columns containing numbers 1-20
matrix_data <- matrix(1:20, nrow = 4)
# Use apply to find the maximum value in each row
row_maxes <- apply(matrix_data, 1, max)
# Display the maximum value from each row
print(row_maxes)

# Exercise 4: Create number pairs using mapply
# Define vectors for start and end points
starts <- 1:3
ends <- 4:6
# Use mapply to create formatted pairs of numbers
number_pairs <- mapply(function(x, y) paste(x, "and", y),
                      starts, ends)
# Display the resulting pairs
print(number_pairs)
```

## Tips and Best Practices

1. **Pre-allocate vectors** in loops for better performance
   - Always create vectors with known size before the loop
   - Avoid growing vectors inside loops
   
2. **Use meaningful iterator names** (not just `i`, `j`, `k`)
   - Choose names that describe what they represent
   - Makes code more readable and maintainable
   
3. **Consider readability** when choosing between loops and apply functions
   - Use loops for complex operations or when code clarity is priority
   - Use apply functions for simpler operations on data structures
   
4. **Test your code** with small examples first
   - Verify logic with simple cases
   - Scale up to larger datasets after confirming correctness
   
5. **Comment your code** to explain complex operations
   - Document the purpose of each major code block
   - Explain any non-obvious logic or calculations 