---
title: "Introduction to R"
author: "Eren Ada"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Getting Started with R

R is a powerful programming language designed for statistical computing and data analysis. This tutorial will cover the fundamental concepts you need to get started with R programming.

### Using R as a Calculator

R can be used as a simple calculator. Type expressions directly in the console:

```{r calculator}
# Basic arithmetic operations
2 + 2       # Addition
10 - 5      # Subtraction
4 * 3       # Multiplication
15 / 3      # Division
2^3         # Exponentiation

# Let's print the results with labels
print("Addition: 2 + 2 =")
print(2 + 2)
print("Exponentiation: 2^3 =")
print(2^3)
```

### Objects and Variables

In R, you can store values in objects using the assignment operator `<-` (or `=`):

```{r objects}
# Creating objects
x <- 10
my_number = 42
text <- "Hello, R!"

# Print objects
print("Value of x:")
print(x)
print("Value of my_number:")
print(my_number)    # Just typing the object name will print it
print("Value of text:")
print(text)
```

### Data Types in R

R has several basic data types:

```{r data-types}
# Numeric
age <- 25

# Character (string)
name <- "John"

# Logical (boolean)
is_student <- TRUE

# Check the type of an object
print("Type of 'age':")
print(class(age))      # "numeric"
print("Type of 'name':")
print(class(name))     # "character"
print("Type of 'is_student':")
print(class(is_student))   # "logical"
```

### Vectors

Vectors are one-dimensional arrays that can hold data of the same type:

```{r vectors}
# Create a vector using c()
numbers <- c(1, 2, 3, 4, 5)
fruits <- c("apple", "banana", "orange")

# Vector operations
numbers_plus_2 <- numbers + 2     # Adds 2 to each element
numbers_times_3 <- numbers * 3     # Multiplies each element by 3

# Vector indexing (starts at 1, not 0)
first_number <- numbers[1]      # First element
second_fruit <- fruits[2]       # Second element

# Print results
print("Original numbers vector:")
print(numbers)
print("Numbers + 2:")
print(numbers_plus_2)
print("Numbers * 3:")
print(numbers_times_3)
print("First number:")
print(first_number)
print("Second fruit:")
print(second_fruit)
```

### Saving Objects

There are several ways to save R objects for later use:

```{r saving-objects}
# Create some objects to save
numbers <- c(1, 2, 3, 4, 5)
fruits <- c("apple", "banana", "orange")

# Save a single object to a .RData file
save(numbers, file = "my_numbers.RData")

# Save multiple objects
save(numbers, fruits, file = "my_objects.RData")

# Save entire workspace
save.image(file = "workspace.RData")

# Load saved objects
load("my_numbers.RData")
print("Loaded numbers:")
print(numbers)
```

### Working with Files

You can write data to and read data from files:

```{r files}
# Create a data frame to write
df <- data.frame(
  numbers = 1:5,
  letters = LETTERS[1:5]
)

# Write a CSV file
write.csv(df, file = "example.csv", row.names = FALSE)

# Read the CSV file back
data <- read.csv("example.csv")
print("Data read from CSV:")
print(data)
```

### Getting Help

R has excellent built-in documentation. Here are some examples of how to access help:

```{r help, eval=FALSE}
# Get help on a function
?mean
help(sum)

# See examples
example(mean)
```

## Practice Exercises

Try these exercises to test your understanding:

### Exercise 1: Create and Manipulate a Vector
```{r exercise1}
# Create a numeric vector with numbers 1 through 10
my_vector <- 1:10

# Calculate the mean
vector_mean <- mean(my_vector)

# Print results
print("My vector:")
print(my_vector)
print("Mean of my vector:")
print(vector_mean)
```

### Exercise 2: Working with Different Data Types
```{r exercise2}
# Create a character vector with three colors
colors <- c("red", "blue", "green")

# Create a list combining numeric and character vectors
my_list <- list(
  numbers = my_vector,
  colors = colors
)

# Print the list
print("My list:")
print(my_list)
```

## Next Steps

After mastering these basics, you can move on to:
- Data frames and tibbles
- Basic plotting with base R
- Installing and using packages
- Basic statistical functions
- Control structures (if/else, loops)

```{r cleanup, include=FALSE}
# Clean up files created during the tutorial
file.remove("my_numbers.RData")
file.remove("my_objects.RData")
file.remove("workspace.RData")
file.remove("example.csv")
``` 